Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Added a placeholder appendix.
merging I guess

@@

\paragraph{$\offerkill$ implementation}

@@

We used this approach to measure the GPU availability required for
three types of rendering: \emph{light}, whereby desktop icons are
smoothly emphasised under the mouse pointer; \emph{medium}, whereby
window dragging over the desktop is smoothly animated; and
\emph{heavy}, which requires smooth animation of a WebGL shader in a
browser.  For \emph{heavy} we used the Paper
Planes\footnote{\url{https://paperplanes.world}}
and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
shaders running in a recent version of Firefox.

@@

Recall that our mega kernel-based implementation merges the code for a
cooperative kernel with that of the non-cooperative task to be
co-scheduled.  This blow-up in code size can have a negative impact on
compilation, e.g.\ due to higher register pressure, reducing the
occupancy for the merged kernel---the maximum number of workgroups
that the GPU driver will schedule.  This is an artifact of our
proof-of-concept implementation strategy, and would not be a problem
if our cooperative scheduling approach was implemented inside the GPU
driver.  We thus launch both the original and cooperative versions of
a kernel with the reduced occupancy bound in order to meaningfully
compare execution times.

@@

Cooperative scheduling enables to run a non-cooperative kernel alongside
a cooperative one, we focus here on the performance impact of this
multitasking and on the scheduling responsiveness. The performance is
measured with respect to the performance of each kernel run in
sequence. Responsiveness, i.e., how fast a non-cooperative kernel can be
scheduled, depends not only on the scheduler implementation, but also on
how often the cooperative kernel offer workgroups to be killed.

% In order to simulate non-persistent workloads alongside kernels of
% blocking algorithms, we use a matrix multiplication kernel where
% matrix dimensions are tuned to trigger the relevant execution
% duration.

\paragraph{Experimental setup}
For a given cooperative kernel and its data set, we make it run while
regularly asking to schedule a non-cooperative kernel that simulates
graphics rendering, according to the classes listed in
\mytab~\ref{tab:noncooperativeworkload}. In practice, we use a matrix
multiplication kernel as the non-cooperative workload, with matrix
dimensions tailored for each GPU to reach the appropriate execution
duration.  In separate runs, we ask the scheduler to launch this
non-cooperative kernels with different number of workgroups: either 1, a
quarter, a half or all but one workgroups among the ones available for
the cooperative kernel. Moreover, for graph algorithms we try both the
regular and the query barrier.

Our experiments span 13 pairs of cooperative kernels and data sets, 3
classes of non-cooperative kernel workloads and frequency, 4 quantities
of workgroups claimed for the non-cooperative kernel and 2 variations of
resizing barriers for graph algorithms, leading to 288
configurations. Each of this configuration is run 10 times on each
GPU. For each run, we record the execution time of the cooperative
kernel. For each scheduling of the non-cooperative kernel during the
run, we also record the \emph{gather time} needed by the scheduler to
collect enough workgroups to launch the non-cooperative kernel, and the
non-cooperative kernel execution time.

Whenever a cooperative or non-cooperative kernel terminates, we verify
that it has really performed its computation by comparing its results
against the expected one. We encountered two kind of issues during the
experiments: some configurations would lead to a system freeze or crash
although they run fine on other GPUs. Although we don't claim that our
programs are all correct, this issue seems to indicate a bug in the
OpenCL compiler or driver of the problematic GPU. We also have some
cooperative kernels that returned invalid results when the
non-cooperative kernel takes all but one of the cooperative kernel
workgroups, we decided to ignore these results.

\paragraph{Results}  \ADComment{Discuss the results.}
explain horizontal lines

missing results for all but one in bfs rmat22 since errors

also, high gather time corresponds to applications and data sets where
resize/kill is triggered less often.

Each configuration was run 10 times on each GPU.

>>>>

\paragraph{Dynamically managing workgroups}

@@

In the case where a cooperative kernel $K_c$ is executing and there
are $W$ avaiable workgroups, the scheduler may decide to give $K_c$
$W_k <= W$ additional workgroups via an $\offerfork$ command issued by
$K_c$. To do this, scheduler signals $K_c$ through the $K_c$'s
scheduler context communication channel to indicate the number of
workgroups that are allowed to join at the next $\offerfork$ command.
Then, when a workgroup $w$ of $K_c$ executes $\offerfork$, $w$
observes $W_k$, and takes $W_k$ workgroups from the pool of available
workgroups, provides them with transmit variables, target, and updated
kernel context.
\TSComment{this is sloppy}

\paragraph{Efficient implementation}

The interesting interactions with respect to performance and safety
occur in the interaction of the scheduler and cooperative kernel. We
discuss these details now.

At the core of the scheduler is the workgroup pool, which maps each
workgroup to a status and keeps track of the number of workgroups in
each status. The workgroup pool can change in two ways: (1)
\emph{assigning tasks}, where workgroups are changed from available to
computing a kernel (2) \emph{becoming available}, where a workgroup
exits a kernel computation (either through completion or $\offerkill$)
and changes its status from computing a kernel to available.

\emph{Assigning tasks} occurs either when the scheduler is launches a
new kernel, or when a workgroup calls $\offerfork$ and observes that
there are available resources. This creates a potential conflict
between forking workgroups as well as the scheduler, as they may try
to assign the same available workgroup to different tasks
simultaneously. To fix this conflict, we introduce an
\emph{assignment} mutex in the scheduler context.

When the scheduler determines a new kernel has been launched, it spins
waiting to acquire the assignment lock. Once acquired, it determines
how to obtain the required number of workgroups (possibly by
signalling a cooperative kernel to enable $\offkill$). The scheduler
than waits until the required number of workgroups become available.
Because it holds the assignment mutex, the scheduler knows that the
number of available workgroups in the workgroup pool can only
increase.

When $\offerfork$ is executed, the calling workgroup has a single
attempt to acquire the assignment mutex.  If it fails, the call
returns without a fork. This provides a quick return of $\offerfork$
when another actor is doing an assigning task. If the assignment mutex
is acquired, then the calling workgroup can check the scheduler
context to see if its allowed to fork by the scheduler and safely
assign available workgroups from the workgroup pool in isolation from
other actors that which to assign tasks. \TSComment{it also updates
  the kernel context, and sends transmit variables while holding this
  lock}

As opposed to assignment tasks, availability tasks can be performed in
a lock-free manner without conflict. This is because a workgroup only
modifies its own state in the task pool and can atomically increment
and decrement the required counters. Availability tasks are also able
to be performed lock-free with assignment tasks (although accesses to
the workgroup state and counters must be performed
atomically). Availability tasks will only increase the number of
available workgroups, and do not conflict with assignment tasks
sending available workgroups to kernels.

Although assignment tasks to the workgroup pool can be performed
concurrently, concurrent calls to $\offerkill$ must be protected
because workgroups that exit (a) modify the kernel context which
causes a conflict and (b) the semantics of $\offerkill$ guarantee that
only the workgroup with the highest id can exit. Rather than introduce
a new mutex to protect $\offerkill$, we instead use the number of
active groups from the kernel context.  The call to $\offerkill$
begins by checking the calling workgroup id against the number of
active groups for the kernel. If the calling workgroup is the last
workgroup, it enters and is then able to check if it can be
killed. Otherwise it immediately returns. If a workgroup determines
it should be killed, it updates

From this implementation, it falls out that $\offerkill$ and
$\offerfork$ from the same workgroup can never succeed concurrently.
If $\offerkill$ terminates the calling workgroup, then the scheduler
is holding the assignment lock and will continue to hold it until 
the call to $\offerkill$ finishes. Thus, any call to $\offerfork$
during this time will not result in forking workgroups.

\paragraph{Example}

@@

We used this approach to measure the GPU availability required for
three types of rendering: \emph{light}, whereby desktop icons are
smoothly emphasised under the mouse pointer; \emph{medium}, whereby
window dragging over the desktop is smoothly animated; and
\emph{heavy}, which requires smooth animation of a WebGL shader in a
browser.  For \emph{heavy} we used the Paper
Planes\footnote{\url{https://www.chromeexperiments.com/experiment/paper-planes}}
and Kaleidoscope\footnote{\url{http://colordodge.com/Kaleidoscope/}}
shaders running in a recent version of Firefox.

@@

Recall that our mega kernel-based implementation merges the code for a
cooperative kernel with that of the non-cooperative task to be
co-scheduled.  This blow-up in code size can have a negative impact on
compiliation, e.g.\ due to higher register pressure, reducing the
occupancy for the merged kernel---the maximum number of workgroups
that the GPU driver will schedule.  This is an artifact of our
proof-of-concept implementation strategy, and would not be a problem
if our cooperative scheduling approach was implemented inside the GPU
driver.  We thus launch both the original and cooperative versions of
a kernel with the reduced occupancy bound in order to meaningfully
compare execution times.

@@

\ADComment{Preamble}

\HEComment{Below, just random notes so far}

Cooperative scheduling enables to concurrently run non-persistent
kernels alongside blocking ones. In terms of performance, we are
interested to see whether the variations of workgroups for the blocking
kernel.

Performance of cooperative scheduling corresponds to execution time of
cooperative kernels compared to the serial

In order to simulate non-persistent workloads alongside kernels of
blocking algorithms, we use a matrix multiplication kernel where
matrix dimensions are tuned to trigger the relevant execution
duration.

we compare :

time to execute original kernel

time to execute non-persistent tasks

time to execute both in cooperation

also, response time for non-persistent

use matmult to represent system workload, ref to previous section

\paragraph{Experimental setup}  \ADComment{Describe the data we are gathering.}

\paragraph{Results}  \ADComment{Discuss the results.}
